Computing in the Field of Rationals

### 2012-06-19 rational_ext.tex
_

@@

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o+ = \multimap^{+}
%! -o* = \multimap^{\times}
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

\title{Computing in the Field of Rationals}
%% \authorinfo{Roshan P. James}
%%            {Indiana University}
%%            {rpjames@indiana.edu}
%% \authorinfo{Amr Sabry}
%%            {Indiana University}
%%            {sabry@indiana.edu}
%% \maketitle

@@

  Previous work on \emph{information effects} established the notion
  of information preserving computation
  \cite{infeffects,rc2011,rc2012} and the core calculus for computing
  with the isomorphisms of finite types, {{langRev}}. Given the
  first-order nature of {{langRev}} however it was unclear how one may
  express first-class functions.  Andrej Filinski's masters thesis
  \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
  symmetry underlying computation in the \lcal. Filinski showed that
  values are dual to continuations and that functions are dual to
  delimited continuations.

  We present a computational model whose types are \emph{the field of
    rational numbers}. This computational model is derived
  systemmatically from the {{langRev}} by considering a symmetric
  notion of duality. Unlike the \lcal which shows a classical demorgan
  dulaity, here we have two axis of dualization and hence two
  dualities -- an additive duality, namely negative types and a
  multiplicative duality, namely fractional types.

@@

Andrej Filinksi's masters thesis
\cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
underlying computation is one embraces the symmetry hidden in
continuations. Filinski showed that values are dual to continuations,
functions are dual to delimited continuations. Filinksi also shows a
de-morgan style duality, by treating value pairs as the dual of
continuation sums and value sums as dual to continuation pairs.

@@

\paragraph*{Contributions and Outline.} 
To summarize, in a computational framework that guarantees that information
is preserved, negative and fractional types provide fascinating mechanisms in
which computations can be sliced and diced, decomposed and recomposed, run
forwards and backwards, in arbitrary ways. The remainder of the paper
formalizes these informal observations. Specifically our main contributions
are:
\begin{itemize}
\item We extend {{langRev}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} (reviewed in Sec.~\ref{sec:pi}) with a notion
  of negative types, that satisfies the isomorphism {{a + (-a) <-> 0}}. The
  semantics of this extension is expressed by having a \emph{dual} evaluator
  that reverses the flow of execution for negative
  values. (Sec.~\ref{sec:neg})
\item We independently extend {{langRev}} with a notion of fractional types,
  that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
  extension is expressed by introducing logic variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types. (Sec.~\ref{sec:frac})
\item We combine the above two extensions into a language, which we
  call {{langRevEE}}, whose type system allows any rational number to
  be used as a type. Moreover the types satisfy the same familiar and
  intuitive isomorphisms that are satisfied in the mathematical field
  of rational numbers. (Sec.~\ref{sec:rat})
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every
  programmer. (Sec.~\ref{sec:prog})
\item We relate our notions of negative and fractional types to previous work
  on continuations. Briefly, we argue that conventional continuations
  conflate negative and fractional components. This observation allows us to
  relate two apparently unrelated lines of work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relating continuations to
  negative types and the second~\cite{Bernardi:2010:CSL:1749618.1749689}
  relating continuations to the fractional types of the Lambek-Grishin
  calculus. (Sec.~\ref{sec:related})
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Core Reversible Language: {{langRev}} }
\label{sec:pi}

We review our reversible language {{langRev}}: the presentation in this
section differs from the one in our previous paper~\cite{infeffects} in two
aspects. First, we add the empty type {{0}} which is necessary to express the
additive duality. Second, instead of explaining evaluation using a natural
semantics, we give a small-step operational semantics that is more
appropriate for the connections with continuations explored in this paper.

@@

The combinators of {{langRev}} are witnesses to the following type
isomorphisms: 
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%zeroe :&  0 + b <-> b &: zeroi
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist0 :& 0 * b <-> 0 &: factor0
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces one or two combinators that witness
the isomorphism in the middle. Collectively the isomorphisms state that the
structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
multiplication distributes over addition. The isomorphisms are extended to
form a congruence relation by adding the following constructors that witness
equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

To summarize, the syntax of {{langRev}} is given as follows.

\begin{definition}{(Syntax of {{langRev}})}
\label{def:langRev}
We collect our types, values, and combinators, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% comb.~types, t ::= b <-> b
% iso ::= zeroe | zeroi 
%     &|& swap+ | assocl+ | assocr+ 
%     &|& identl* | identr* 
%     &|& swap* | assocl* | assocr* 
%     &|& dist0 | factor0 | dist | factor 
% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

@@

%% In fact, when talking about {{langRev}} circuits, we will often use the
%% words {{value}} and {{particle}} interchangeably.

@@

The semantics of the primitive combinators is given by the following
single-step reductions. Since there are no values of type {{0}}, the rules
omit the impossible cases:
\begin{scriptsize}
%subcode{opsem} include main
%! columnStyle = rlcl
% swap+ & (left v) &|-->& right v
% swap+ & (right v) &|-->& left v 
% assocl+ & (left v1) &|-->& left (left v1)
% assocl+ & (right (left v2)) &|-->& left (right v2)
% assocl+ & (right (right v3)) &|-->& right v3 
% assocr+ & (left (left v1)) &|-->& left v1
% assocr+ & (left (right v2)) &|-->& right (left v2)
% assocr+ & (right v3) &|-->& right (right v3)
% identl* & ((), v) &|-->& v 
% identr* & v &|-->& ((), v) 
% swap* & (v1, v2) &|-->& (v2, v1) 
% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
% dist & (left v1, v3) &|-->& left (v1, v3)
% dist & (right v2, v3) &|-->& right (v2, v3)
% factor & (left (v1, v3)) &|-->& (left v1, v3) 
% factor & (right (v2, v3)) &|-->& (right v2, v3)   
\end{scriptsize}
The reductions for the primitive isomorphisms above are exactly the same as
have been presented before~\cite{infeffects}. The reductions for the closure
combinators are however presented in a small-step operational style using the
following definitions of evaluation contexts and machine states:

@@

\begin{proposition}[Logical Reversibility]
\label{prop:logrev}
{{<c,v,[]> |--> [c,v',[]]}} iff 
{{<c{dagger},v',[]> |--> [c{dagger},v,[]]}} 
\end{proposition}

@@

% \begin{proposition}[Groupoid]
% \label{prop:groupoid}
% {{langRev}} is a groupoid. 
% \end{proposition}

% \begin{proposition}
% \label{prop:category}
% {{langRev}} is a dagger symmetric monoidal category. 
% \end{proposition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{First class functions from the permutation space.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Negative and Fractional Types}
\label{sec:neg}
\label{sec:frac}

This section introduces the syntax and graphical languages for
negatives and fractionals.  The general outline is as follows:
\footnote{We refer the reader to Selinger's excellent survey
  article of monoidal
  categories~\cite{springerlink:10.1007/978-3-642-12821-94} for the precise
  definitions.}
\begin{itemize}
\item As established in our earlier work~\cite{rc2011,infeffects}, the
  underlying categorical semantics of our core reversible language
  {{langRev}} is based on \emph{two} distinct symmetric monoidal structures,
  one with {{+}} as the monoidal tensor, and one with {{*}} as the monoidal
  tensor.
\item We extend each underlying symmetric monoidal structure to a
  \emph{compact closed} structure by adding a dual for each object and two
  special morphisms traditionally called {{eta}} and {{eps}}. 
\end{itemize}
The extended language is referred to as {{langRevEE}}.  After
presenting the extension at the syntactic level, we discuss the
categorical semantics informally via the graphical language. The
operational semantics is presented in Sec.~\ref{sec:rat}.

As will be detailed in the remainder of this section, the compact closed
structure provides several properties of interest: (i) morphisms or wires are
allowed to run from right to left, (ii) the structure admits a trace that,
depending on the situation, can be used to implement various notions of loops
and
recursion~\cite{joyal1996traced,Hasegawa:2009:TMC:1552068.1552069,Hasegawa:1997:RCS:645893.671607},
(iii) the structure includes an isomorphism showing that the dual operator is
an involution, and (iv) the structure is equipped with objects representing
higher-order functions and a morphism \textit{eval} that applies these
functional objects. Interestingly each monoidal structure provides the same
ingredients, resulting in two dualities, two traces, two involutions, and two
notions of higher-order functions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Syntax and Types}

Describing the syntax and types of an extension to {{langRev}} with additive
and multiplicative duality is fairly straightforward. Basically, for the
additive case, we extend the language with negative types denoted {{-b}},
negative values denoted {{-v}}, and two isomorphisms {{eta+}} and
{{eps+}}. Similarly, for the multiplicative case, we extend the language with
fractional types denoted {{1/b}}, fractional values denoted {{1/v}}, and two
isomorphisms {{eta*}} and {{eps*}}.

%subcode{bnf} include main
% Value Types, b = ... | -b | 1/b
% Values, v = ... | -v | 1/v
%
% Isomorphisms, iso &=& ... | eta+ | eps+ | eta* | eps*

For convenience, we sometimes use the notations {{b1 - b2}} and
{{b1/b2}} to indicate the types {{b1 + (-b2)}} and {{b1 * (1/b2)}}
respectively.  The types of the new constructs are:
\vspace{-15pt}
\begin{multicols}{2}  
%subcode{opsem} include main
% eta+ &: 0 <-> (-b) + b :& eps+
% eta* &: 1 <-> (1/b) * b :& eps*

%subcode{proof} include main
%@ |- v : b
%@@ |- -v : -b
%
%@ |- v : b
%@@ |- 1/v : 1/b
\end{multicols}

For the graphical language, we visually represent {{eta+}}, {{eps+}},
{{eta*}}, and {{eps*}} as U-shaped connectors. On the left below are {{eta+}}
and {{eta*}} showing the maps from {{0}} to {{-b+b}} and {{1}} to {{1/b * b}}. 
On the right are {{eps+}} and {{eps*}} showing the maps from {{-b+b}} to~0 and 
{{1/b * b}} to~1. Even though the diagrams below show the {{0}} and {{1}} 
wires for completeness, later diagrams will always drop them in contexts 
where they can be implicitly introduced and eliminated. 
\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/eta.pdf}
\end{center}

\begin{center}
 \includegraphics{diagrams/eps.pdf}
\end{center}
\end{multicols}
\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/eta_times.pdf}
\end{center}

@@

The usual interpretation of the type {{b1+b2}} is that we either have an
appropriately tagged value of type {{b1}} or an appropriately tagged value of
type {{b2}}. This interpretation is maintained in the presence of {{eta+}}
and {{eps+}} in the following sense: a value of type {{right v : -b + b}}
flowing into an {{eps+}} on the lower wire switches to a value 
{{left (-v): -b + b}} that flows backwards on the upper wire. The inversion of
direction is captured by the negative sign on the value.

\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eps_plus2.pdf}
}
\end{center}  
\end{multicols}
As an example, the diagram representing the first isomorphism in the
introduction is:
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/algebra1.pdf}
}
\end{center}

Similarly, in the usual interpretation of {{b1*b2}} we have both a value of
type {{b1}} and a value of type {{b2}}. This interpretation is maintained in
the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is to be
viewed as a fission point for a value of type {{b}} and its multiplicative
inverse {{1/b}}. Operationally, this corresponds to the creation of two
values {{alpha}} and {{1/alpha}} where {{alpha}} is a fresh logic
variable. The operator {{eps*}} then becomes a unification site for these
logic variables:
\begin{multicols}{2}
\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eta_times1.pdf}
}
\end{center}

@@

It is useful to contrast the categorical structure of {{langRevEE}},
which has two distinct \emph{compact closed} structures, to the
well-known and closed related \emph{*-autonomous} categories that form
a model for linear logic. The distnguishing feature that seperates
*-autonomous structures from compact closed structures is that the
dualizing action in *-autonomous categories does not preserve the
tensor. In {{langRev}}, as we will see , we can construct the
isomorphism {{-(b1+b2)<-> (-b1)+(-b2)}}, i.e. the tensor {{+}} has
been preserved by the negation.  Whereas in linear logic we have that
$(a\oplus b)^\bot$ is $a^\bot \parr b^\bot$, i.e. the connective
$\oplus$ is not preserved.

In linear logic terms `additive pairs' and `multiplicative sums' are
used to indicate this notion that some connectives has both additive
and multiplicative properties. The usual intuitive interpretations for
these involve suggestions like one has a dollar to spend and the both
cigarettes and soda are available for purchase, but the fact that you
have only a dollar to spend affords you only one of these. The
intuitions talk about a sort of dependacy underlying the fact that
both cigarettes and soda are available but you can have only one.

In {{langRevEE}}, operators {{eta}} and {{eps}} capture a similar
notion.The pair of values {{v}} and {{1/v}} created by {{eta*}} are
not fully independant in the way one would think of normal pairs. The
value of one, affects the value of the other. Similarly, the {{v}} and
{{-v}} created by {{eta+}} do really present a strict choice of
{{left}} or {{right}} in the sense that usual sums do, but instead can
have a \emph{both} {{left}} and {{right}} values at different points
during the computation. However, in our opinion, the philisophical
difference the additiveness of the additive pairs and the
multiplicativesness of the multiplicative pair aren't a property of
the connective between the values, but a property intrinsic to the
values themselves. Hence {{v}} and {{1/v}} share secretely a
relationship that their combinting {{*}} constructor does not and need
not know about.

@@

\paragraph*{Duality preserves the monoidal tensor. }
As with compact closed categories, the dual on the objects distributes
over the tensor. In terms of {{langRevEE}} we have that {{-(b1+b2)}}
can be mapped to {{(-b1)+(-b2)}} and that {{1/(b1*b2)}} can be mapped
to {{(1/b1)*(1/b2)}}. The isomorphism {{-(b1 + b2) <-> (-b1) + (-b2)}}
can be realized as follows:
\begin{center}
  \includegraphics{diagrams/dist_neg_plus.pdf}
\end{center}

@@

\paragraph*{Duality is a functor.}
Duality in {{langRevEE}} can map objects to their duals and morphisms
to act on dual objects. In other {{c : b1 <-> b2}} to
{{neg~c:-b1<->-b2}} in the additive monoid and to
{{inv~c:1/b1<->1/b2}} in the multiplicative monoid. 
%% Any operation on types can be applied to the negative versions of these
%% types. 
The idea is simply to reverse the flow of values and use the
adjoint of the operation:
\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/neg_lift.pdf}
\end{center}

@@

This construction relies on the fact that every {{langRevEE}} morphism
has an adjoint.  The {{inv}} construction is similar.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Functions and Delimited Continuations}
\label{sub:hof}

%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Additional Constructions}
\label{sec:specific-constructions}

% \paragraph*{Unification.}
% A computational interpretation of this is that {{langRevEE}} acts like
% a logic programming system and {{eta*}} acts as the site of creation
% of a new typed logic variable of type {{b}} and its corresponding dual
% of type {{1/b}}. These two values are free to move independently
% through a circuit, but any operations on value affect what operations
% are possible on the other since they are both related by an underlying
% logic variable. Dually, {{eps*}} acts as a unification site for values
% flowing along both branches. If the values are not exactly duals, then
% the system fails to unify and hence the program is stuck. If the
% values are exactly duals, then unification succeeds and {{eta*}}
% returns {{1}}.

>>>>

\newcommand{\asterix}[0]{*}

@@

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{\asterix}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o+ = \multimap^{+}
%! -o* = \multimap^{\times}
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! Pi = \Pi
%! Pi0 = \Pi^{o}
%! PiEE* = \Pi^{\eta\epsilon}_{*}
%! PiEE+ = \Pi^{\eta\epsilon}_{+}
%! PiEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

\title{Computing in the Field of Rationals}

\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}

\maketitle

@@

  Andrzej Filinski's masters thesis
  \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
  symmetry underlying computation in the \lcal. Filinski showed that
  values are dual to continuations and that functions are dual to
  delimited continuations, thereby suggesting that
  values-and-functions and continuations-and-delimited-continuations
  are mirror images of the same phenomena.

  Previous work on \emph{information effects} established the notion
  of information preserving computation
  \cite{infeffects,rc2011,rc2012} and the core calculus for computing
  with the isomorphisms of finite types, {{Pi}}. Given the
  first-order nature of {{Pi}} however it was unclear how one may
  express first-class functions and duality in the sense of Filinski.

  We present a computational model whose types are \emph{the field of
    rational numbers}. This computational model is derived
  systemmatically from the {{Pi}} by considering a symmetric
  notion of duality. Unlike the \lcal which shows a classical demorgan
  duality, here we have two axis of dualization and hence two
  dualities -- an additive duality, namely negative types and a
  multiplicative duality, namely fractional types.

@@

Andrzej Filinksi's masters thesis
\cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
underlying computation in the \lcal. Filinski showed that values are
dual to continuations, functions are dual to delimited
continuations. Filinksi also shows a De Morgan style duality, by
treating value pairs as the dual of continuation sums and value sums
as dual to continuation pairs.

@@

\paragraph*{Contributions and Outline.} 
To summarize, in a computational framework that guarantees that information
is preserved, negative and fractional types provide fascinating mechanisms in
which computations can be sliced and diced, decomposed and recomposed, run
forwards and backwards, in arbitrary ways. The remainder of the paper
formalizes these informal observations. Specifically our main contributions
are:
\begin{itemize}
\item We extend {{Pi}} our reversible programming language of type
  isomorphisms~\cite{rc2011,infeffects} (reviewed in Sec.~\ref{sec:pi}) with a notion
  of negative types, that satisfies the isomorphism {{a + (-a) <-> 0}}. The
  semantics of this extension is expressed by having a \emph{dual} evaluator
  that reverses the flow of execution for negative
  values. (Sec.~\ref{sec:neg})
\item We independently extend {{Pi}} with a notion of fractional types,
  that satisfies the isomorphism {{a * (1/a) <-> 1}}. The semantics of this
  extension is expressed by introducing logic variables and a unification
  mechanism to model and resolve the constraints introduced by the fractional
  types. (Sec.~\ref{sec:frac})
\item We combine the above two extensions into a language, which we
  call {{PiEE}}, whose type system allows any rational number to
  be used as a type. Moreover the types satisfy the same familiar and
  intuitive isomorphisms that are satisfied in the mathematical field
  of rational numbers. (Sec.~\ref{sec:rat})
\item We develop programming intuition and argue that negative and fractional
  types ought to be part of the vocabulary of every
  programmer. (Sec.~\ref{sec:prog})
\item We relate our notions of negative and fractional types to previous work
  on continuations. Briefly, we argue that conventional continuations
  conflate negative and fractional components. This observation allows us to
  relate two apparently unrelated lines of work: the first pioneered by
  Filinski~\cite{Filinski:1989:DCI:648332.755574} relating continuations to
  negative types and the second~\cite{Bernardi:2010:CSL:1749618.1749689}
  relating continuations to the fractional types of the Lambek-Grishin
  calculus. (Sec.~\ref{sec:related})
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Core Reversible Language: {{Pi}} }
\label{sec:pi}

We review our reversible language {{Pi}}: the presentation in this
section differs from the one in our previous paper~\cite{infeffects} in two
aspects. First, we add the empty type {{0}} which is necessary to express the
additive duality. Second, instead of explaining evaluation using a natural
semantics, we give a small-step operational semantics that is more
appropriate for the connections with continuations explored in this paper.

@@

The combinators of {{Pi}} are witnesses to the following type
isomorphisms: 
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%zeroe :&  0 + b <-> b &: zeroi
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist0 :& 0 * b <-> 0 &: factor0
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces one or two combinators that witness
the isomorphism in the middle. Collectively the isomorphisms state that the
structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
multiplication distributes over addition. The isomorphisms are extended to
form a congruence relation by adding the following constructors that witness
equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

To summarize, the syntax of {{Pi}} is given as follows.

\begin{definition}{(Syntax of {{Pi}})}
\label{def:Pi}
We collect our types, values, and combinators, to get the full language
definition.
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b+b | b*b 
% values, v ::= () | left v | right v | (v,v) 
%
% comb.~types, t ::= b <-> b
% iso ::= zeroe | zeroi 
%     &|& swap+ | assocl+ | assocr+ 
%     &|& identl* | identr* 
%     &|& swap* | assocl* | assocr* 
%     &|& dist0 | factor0 | dist | factor 
% comb., c ::= iso | id | sym c | c (;) c | c (+) c | c (*) c 
\end{definition}

@@

%% In fact, when talking about {{Pi}} circuits, we will often use the
%% words {{value}} and {{particle}} interchangeably.

@@

The semantics of the primitive combinators is given by the following
single-step reductions. Since there are no values of type {{0}}, the rules
omit the impossible cases:
\begin{scriptsize}
%subcode{opsem} include main
%! columnStyle = rlcl
% swap+ & (left v) &|-->& right v
% swap+ & (right v) &|-->& left v 
% assocl+ & (left v1) &|-->& left (left v1)
% assocl+ & (right (left v2)) &|-->& left (right v2)
% assocl+ & (right (right v3)) &|-->& right v3 
% assocr+ & (left (left v1)) &|-->& left v1
% assocr+ & (left (right v2)) &|-->& right (left v2)
% assocr+ & (right v3) &|-->& right (right v3)
% identl* & ((), v) &|-->& v 
% identr* & v &|-->& ((), v) 
% swap* & (v1, v2) &|-->& (v2, v1) 
% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
% dist & (left v1, v3) &|-->& left (v1, v3)
% dist & (right v2, v3) &|-->& right (v2, v3)
% factor & (left (v1, v3)) &|-->& (left v1, v3) 
% factor & (right (v2, v3)) &|-->& (right v2, v3)

%subcode{proof} include main
%@ ~
%@@ id v |--> v 
%
%@ c{dagger} v1 |--> v2
%@@ (sym c) v1 |--> v2
%
%@ c1  v1 |--> v
%@ c2  v |--> v2
%@@ (c1(;)c2)  v1 |--> v2
%---
%@ c1  v1 |--> v2
%@@ (c1 (+) c2)  (left v1) |--> left v2
%
%@ c2  v1 |--> v2
%@@ (c1 (+) c2)  (right v1) |--> right v2
%---
%@ c1  v1 |--> v3
%@ c2  v2 |--> v4
%@@ (c1 (*) c2)  (v1, v2) |--> (v3, v4)
\end{scriptsize}

\begin{proposition}[Type Safety]

\end{proposition}

\begin{proposition}[Strong Normalizing]
\label{prop:termination-pi} 
{{Pi}} computation always terminate.

{{forall. c:b1<->b2, v:b1, exists v':b2.}}  
{{c v |--> v'}}
\end{proposition}

\begin{proposition}[Logical Reversibility]
\label{prop:logrev}
{{c v |--> v'}} iff 
{{ c{dagger} v' |--> v}}
\end{proposition}

\subsection{Small Step Semantics}

The reductions for the primitive isomorphisms above are exactly the same as
have been presented before~\cite{infeffects}. The reductions for the closure
combinators are however presented in a small-step operational style using the
following definitions of evaluation contexts and machine states:

@@

\begin{proposition}[Correspondence]
  Evalution in the small-step evaulator corresponds to evaluation in
  the natural semantics.

  {{c v |--> v'}} iff {{<c, v, []> |-->* [c, v', [] ]}}

\end{proposition}

Consequently the small step semantics is logically reversible and
strong normalizing.

\begin{proposition}[Type Safety]

\end{proposition}

@@

% \begin{proposition}[Groupoid]
% \label{prop:groupoid}
% {{Pi}} is a groupoid. 
% \end{proposition}

% \begin{proposition}
% \label{prop:category}
% {{Pi}} is a dagger symmetric monoidal category. 
% \end{proposition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{First class functions from the permutation space.}

@@

For a traced symmetric monoidal catgeory, the Int-construction of
Joyal, Street and Verity \cite{joyal1996traced} gives us a way of
constructing a compact closed category. This is the obvious first
thing to try for a duality in this setting.

%subcode{proof} include main
%@ c : b1 + b2 <-> b1 + b3
%@@ trace c : b2 <-> b3

%% \begin{center}
%%   \includegraphics{diagrams/thesis/trace_plus.pdf}
%% \end{center}

Show that the int construction does not preserve the {{*}} tensor. We
need a duality that preserves both tensors.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types : {{PiEE*}} }

%subcode{bnf} include main
% Value Types, b = 0 | 1 | b + b | b * b  | 1/b
% Values, v = () | left v | right v | (v, v) | 1/v
%
% Isomorphisms, iso &=& ... | eta* | eps*

\noindent
No division by zero.

\begin{multicols}{2}  
%subcode{opsem} include main
% eta* &: 1 <-> (1/b) * b :& eps*

%subcode{proof} include main
%@ |- v : b
%@@ |- 1/v : 1/b
\end{multicols}

For the graphical language, we visually represent {{eta*}}, and
{{eps*}} as U-shaped connectors. On the left below is {{eta*}} showing
the map and {{1}} to {{1/b * b}}.  On the right is {{eps*}} showing
the map from {{1/b * b}} to~1. Even though the diagrams below show the
{{1}} wires for completeness, later diagrams will always drop them in
contexts where they can be implicitly introduced and eliminated.
\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/eta_times.pdf}
\end{center}

@@

The usual interpretation of {{b1*b2}} we have both a value of type
{{b1}} and a value of type {{b2}}. This interpretation is maintained
in the presence of fractionals. Hence an {{eta* : 1 <-> (1/b) * b}} is
to be viewed as a fission point for a value of type {{b}} and its
multiplicative inverse {{1/b}}. Operationally, this corresponds to the
creation of two values {{alpha}} and {{1/alpha}} where {{alpha}} is a
fresh logic variable. The operator {{eps*}} then becomes a unification
site for these logic variables:

\begin{multicols}{2}
\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eta_times1.pdf}
}
\end{center}

@@

Negative Information and Measurements.

Compact Closed Structure (modulo 0).

Show that ``no division by zero'' is supported by the semantics.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Semantics}

Provide a semantics with the nondeterminism monad and define an
\emph{eval}.

\begin{proposition}[Type Safety]

\end{proposition}

\begin{proposition}[Strong Normalizing]

\end{proposition}

Note that this language is not logically reversible.

\begin{proposition}[Coherence]

\end{proposition}

@@

\paragraph*{Duality preserves the monoidal tensor. }
As with compact closed categories, the dual on the objects distributes
over the tensor. In terms of {{PiEE}} we have that {{-(b1+b2)}}
can be mapped to {{(-b1)+(-b2)}} and that {{1/(b1*b2)}} can be mapped
to {{(1/b1)*(1/b2)}}. The isomorphism {{-(b1 + b2) <-> (-b1) + (-b2)}}
can be realized as follows:
\begin{center}
  \includegraphics{diagrams/dist_neg_plus.pdf}
\end{center}

@@

\paragraph*{Duality is a functor.}
Duality in {{PiEE}} can map objects to their duals and morphisms
to act on dual objects. In other {{c : b1 <-> b2}} to
{{neg~c:-b1<->-b2}} in the additive monoid and to
{{inv~c:1/b1<->1/b2}} in the multiplicative monoid. 
%% Any operation on types can be applied to the negative versions of these
%% types. 
The idea is simply to reverse the flow of values and use the
adjoint of the operation:
\begin{multicols}{2}
\begin{center}
  \includegraphics{diagrams/neg_lift.pdf}
\end{center}

@@

This construction relies on the fact that every {{PiEE}} morphism
has an adjoint.  The {{inv}} construction is similar.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{SAT Solver }
\label{sec:prog}
\label{sec:sat-solver}
